# Prerequisites {#prereq}

## Software needed

1\. Download and install R from: https://cran.r-project.org/  

2\. Highly recommended: download and install RStudio from: https://posit.co/download/rstudio-desktop/

3\. Install packages (only once!)  

After you've installed R and RStudio, copy and paste the code below into the RStudio console and hit enter. 
This will install all of the packages in the list labeled `pkg`.
You only need to do this once, but you may need to update packages in the future (especially if you update your R version).

```{r pkg-load-01, eval = FALSE}
pkg <- c('vegan', 'betapart', 'labdsv', 'ade4', 'ecodist', 'ape', 'picante')
install.packages(pkg)
```

4\. Load packages  

```{r pkg-load-02, eval = FALSE}
sapply(pkg, require, character.only = TRUE)
```
    
```{r pkg-load-03, eval = TRUE, echo = FALSE}
pkg <- c('vegan', 'betapart', 'labdsv', 'ade4', 'ecodist', 'ape', 'picante')
snk <- capture.output(sapply(pkg, require, character.only = TRUE))
```

5\. Load some utility functions  

```{r source-utils, eval = FALSE}
source("https://github.com/mlammens/esa2025/raw/main/R/utils_color.R")
```

## Get the data

```{r get-dataset-gh, eval=FALSE}
u <- 'https://github.com/mlammens/esa2025/raw/main/data/veg.rda'
u <- url(u)
load(u)
```

<!---
## Original materials from template below

You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].
--->
