# Introduction to R and RStudio {#intro-to-RStudio}

## Learning Outcomes

At the end of this session you should be able to:

* Perform basic arithmetic using R/RStudio
* Identify the major components of the RStudio interface
* Create an Rmd file and use it to document a sequence of R functions
* Read in a data set and perform basic data summary procedures in R

## Getting started

For this session, we will be working in RStudio (see section \@ref(prereq) for install directions). 
Alternatively, you can work in a cloud-based tool such as Posit Cloud [https://posit.cloud](https://posit.cloud) or Google Colab [https://colab.research.google.com](https://colab.research.google.com), however we will not be going over how to use these tools in detail today.

### Difference between R and RStudio

In this workshop, we will be working in RStudio.
So what is the difference between R and RStudio?
R is a programming language (specifically a statistical analysis programming language), while RStudio is an **Integrated Development Environment**, or **IDE**. 
You can think of it as a interface with R that aids the analyst.
RStudio offers a number of features, mostly related to visual presentation of information, that make writing and working with R code easier.

### Overall layout

There are four panels in the RStudio interface (though you may only have three open when you first start it), each has valuable information.

* Console / Terminal panel (lower-left)
* Environment / History / Git (upper-right)
* Files / Plots / Packages / Help (lower-right)
* Source / Editor (upper-left)

### File management

Good file management is important in data analysis. We'll start working on this now.

* Make a `ESA_QuantComm` directory / folder
* Make a `data` directory / folder within `ESA_QuantComm`
* Make a `scripts` directory / folder within `ESA_QuantComm`
* In RStudio, set working directory to your new ``ESA_QuantComm`` directory

#### Setting your working directory

Point-and-click method - Use 'Session' > 'Set Working Directory' > 'Choose Directory'.

Using the R Console:

```{r, eval=FALSE}
setwd("/Users/maiellolammens/ESA_QuantComm/")
```

Next, we will make an R Project. The benefit to using R Projects is that it will reside within your project folder and act as "glue" to make it easier to connect your data files to your code files. It also encourages good file management by having related files all in the same location or set of folders. If you use R Projects, you can skip this step of setting a working directory, which won't be needed, as it will use "relative paths" to find your files, located in the same location as the .Rproj file that the next step will create. This can be especially helpful if you work on more than one computer, or are sharing files with collaborators, who won't need to change the working directory anymore.


### Making an R Project

Let's make a new R Project associated with your `ESA_QuantComm` directory. 
To make a new project, got to the upper right-hand side of the RStudio interface, where it says **Project: (None)**.
Click the little downward arrow, select "New Project", then select "Existing Directory" from the window that pops up. 
Use the graphical user interface (GUI) to navigate to the `ESA_QuantComm` directory, then select "Create Project".

### Getting help

* **Help** panel (lower right corner)
* `help.search`

```{r}
help.search("bar plot")
```

#### Challenge

Use the `help.search` function to search for something in statistics that you think should be in R? Did you find anything?

> I know my function, I just want the details. Then use the `?` operator - `?barplot`


## R as a calculator

We can use R just like any other calculator. Execute the commands below in the RStudio **Console** panel. After you write the code, press enter to execute it. **NOTE:** in most cases, *white space* (i.e., extra spaces) is ignored by R. So `3+5` and `3 + 5` result in identical output.

```{r}
3 + 5
```

There's internal control for order of operations (PEM DAS = parentheses, exponents, multiplication, division, addition, subtraction). Compare the lines below, noting the position of the parentheses. 

```{r}
(3 * 5) + 7
```

```{r}
3 * 5 + 7
```

```{r}
3 * (5 + 7)
```

> **COMMON ERROR! - incomplete line of code**   
> In the console, you should see the `>` at the start of each new line, but sometimes something may go wrong and you will see a `+` at the start of the line instead. This means that R is waiting for you to complete the line of code. A very common mistake that causes this is failing to "close" parantheses. Can you find the mistake in the code chunk below?

```
> 3 * (5 + 7
+
+
```

### Internal functions

There are a ton of internal functions, and a lot of add-ons via additional packages. Here are a few functions.

```{r}
sqrt(4)
abs(-5)
round(3.14159265359, 2)
```

Functions are always followed by parentheses. Inside the parentheses is where you put the function **arguments**. You can *explicitly* define arguments (which is good practice) or you can allow them to be defined *implicitly*. For example, the function `round` has two arguments, the number you want to round, `x`, and the number of digits you want to round to, `digits`. Above we used implicit argument assignment. Here's an example of assigning values to these arguments explicitly.

```{r}
round(x = 3.14159265359, digits = 2)
```


## Documenting your work with R Markdown

Using [R Markdown](https://rmarkdown.rstudio.com/), we can integrate descriptive text, R code, and the output from that code, into a seamless document that can be easily **reproduced**.
Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. 
While some of the details of using Markdown (and thus R Markdown) can get tricky, the basics are very easy to pickup.
RStudio even comes with two very useful tools to learn and use Markdown. 
First, go to the `Help` tab and select `Markdown Quick Reference`.
Second, a more detailed reference can be found in the `Help` -> `Cheatsheets` section.

### Markdown vs R Markdown

R Markdown is a tool that allows you to make a simple text document using the Markdown formatting syntax **with** R code and associated output embedded. 

### Starting with R Markdown

1. Go to the new file button (upper left corner of the RStudio interface, or `File` -> `New File`) and choose the R Markdown option. 
2. You may be prompted to install new R packages, let's go ahead and do this if you are prompted.
3. A screen will pop-up requesting information such as Title, Author, and Output Format. 
Give your new doc a title and select **HTML** as the output format.
4. A new file that is pre-populated with a bunch of text will open in the `Source` panel.

We will go over each section briefly:

* Header
* Sections
* Code Chunks (The code chunk syntax is *very* specific and important. As a beginner, you may choose to use the GUI button to make empty chunks - it's the little green box with a **C** in it and a plus sign.)

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. Click **Knit** now to see what happens.

## R script file

Another common file format for documenting your work is an *.R script file.
Script files are primarily filled with only code and minimal *comments* that you can include to explain that code.

> **Important:** within an R file, you can use the `#` sign to add comments. Anything written after the # is **not** interpreted when you run the code. You can print a script file, but it will print as a simple text file without any formatting.

***

#### Challenge

In the Rmd file we created above, create a new R chunk and add some math operations. Now `Knit` the new file. 

***

#### Miscellaneous file management points and other stuff

* Navigating the file path - similar to how you would navigate using the command line / terminal
* Tab completion of file paths
* Tab completion of R commands
* The cheat - `file.choose()`

***

#### Challenge - tab completion

* Try to auto-complete `fil`, what do you find? Hint: you hit the `tab` button to show auto-complete options once you have started typing.
* Use the brief help menu that comes up to find a function that starts with `file`, and describe what you think it does.

***

## Value types in R

There are several different types of values that can be used in a programming language. The ones that are most common in R are:

* **character**: A character string. This can be numbers, letters, symbols, or a combination of these things. This is denoted by putting the value inside quotes ("a", "cat", "Aayush", "89", "2e34g6")
* **double**: A numeric value that could be a decimal (not just whole numbers). Numbers are not inside quotes.
* **int**: A numeric value that can only be a whole number
* **factor**: A value that represents a grouping variable or a treatment level. For example, if we had a study with 4 sites, labeled 1, 2, 3, and 4, we would need to treat those as factors, not as numeric values. Similarly, if I wanted to compare academic success among class years labeled freshmen, sophomore, junior, and senior, I would need to treat those as factors, not as character values.
* **logical**: A value that represents TRUE or FALSE. Numbers can be forced into logical values, but only 0 is assigned FALSE; all other values are assigned TRUE.

There are other types in R, but these are the most important to identify as we get started.


## Variables and objects

There are several basic types of data structures in R.

* **VECTORS**: One-dimensional arrays of numbers, character strings, or logical values (T/F)
* **FACTORS**: One-dimensional arrays of factors (Stop - Let's discuss factors)
* **DATA FRAMES**: Data tables in which the various columns may be of different type
* **MATRICES**: In R, matrices can only be 2-dimensional, anything higher dimension is called an array (see below). Matrix elements are numerical; some functions, like the transpose function t(), only work on matrices
* **ARRAYS**: higher dimensional matrices are called arrays in R
* **LISTS**: lists can contain any type of object as list elements. You can have a list of numbers, a list of matrices, a list of characters, etc., or any combination of the above.

### Functions that are useful for understanding the different types of data structures

These functions will tell you what *kind* of variable you are dealing with, as well as some additional information which may be useful as you advance in your use of R.

```
str()
class()
```

## Practice with variables

Let's define a variable.

```{r}
my_var <- 8
```

And another

```{r}
my_var2 <- 10
```

Work with vars

```{r}
my_var + my_var2
```

Make a new variable

```{r}
my_var_tot <- my_var + my_var2
```

***

#### Challenge

Change the value of my_var2

```{r}
my_var2 <- 3
```

What is the value of `my_var_tot` now? Why?

***

### Make a vector

Let's combine multiple values into a vector of length greater than 1. This introduces us to the one of the most prevalent functions in R, `c()` - i.e., the *combine* function.

```{r}
# Vector of variables
my_vect <- c(my_var, my_var2)

# Numeric vector
v1 <- c(10, 2, 8, 7, 11, 15)

# Char vector
pets <- c("cat", "dog", "rabbit", "pig")
```

Making a vector of numbers in sequence

```{r}
v2 <- 1:10
v3 <- seq(from = 1, to = 10)
```

***

#### Challenge

1. Look up the help for the `seq` function, and use this to make a vector from 0 to 100, by steps of 5.
2. Come up with a way that you would use the `length.out` argument.

***

## Exploring variable elements

You can get specific elements from vectors and other data structures

* Introduction to the square brackets `[]`

```{r}
pets <- c("cat", "dog", "rabbit", "pig", "snake")
pets[1]
```

* Getting a number of elements, in sequence

```{r}
pets[3:4]
```

* Getting a number of elements, not in sequence

```{r}
pets[c(1,4)]
```


## Data frames

### Reading in your own data

One of the most basic things you will need to do in R is read in your own data set. 
You can read in Excel files, simple text files, and even files from Google Sheets (this is a bit more complicated though).
But the easiest type of file to read in is a comma separated values (CSV) file. 
You can save an Excel workbook (or Numbers workbook or Google Sheet) as a CSV file by using the "Save as ..." menu item. 

Let's read in a data set of trait measurements from 56 different plant species using the function `read.csv`.
First, you'll need to download the data set from the workshop site. 
Second, move the file to the `data` sub-folder in your `ESA_QuantComm` folder.
Then, we'll read in the file. **NOTE: you may need to adjust the path to the file.**

```{r}
trait_data <- read.csv("data/tra.csv")
```

Let's have a brief look at the first and last six rows of data.

```{r}
head(trait_data)
tail(trait_data)
```

#### `summary` function

Next we'll use the `summary` function to examine this data set.
`summary` returns many standard statistics.
When doing data exploration, a few things you want to look at are:

* How do the mean and median values within a variable compare?
* Do the min and max values suggest there are outliers?
* Which variables (i.e., columns) are quantitative (numeric) versus categorical (factors or characters)

```{r}
summary(trait_data)
```

#### A (very) brief introduction to navigating a `data.frame`

We will be very brief here. 
Check out 
[this Data Carpentry lesson](http://www.datacarpentry.org/R-ecology/03-data-frames.html) for more information.

* Looking at specific `data.frame` elements. Use *row* and *column* notation.

Here is the 5th row, 3rd column (Site).
**NOTE: We are using square brackets to index the `data.frame` and we *always* use [row, column] notation.**

```{r}
trait_data[5, 3]
```

* Looking at an entire column.

Here are two ways to get the `min_height` column.

First, **note: we leave the row part blank, but still add the comma.**

```{r}
trait_data[ ,9]
```

Second, **use only the variable (column) name. Note the use of the `$` operator**

```{r}
trait_data$min_height
```

* Looking at specific column entry

This is a way to look at the 5th entry in the `min_height` column.

```{r}
trait_data$min_height[5]
```

* Looking at all entries for a given row.

Here's all the entries for the 5th row.
**NOTE: here we leave the column part blank, but still add the comma.**

```{r}
trait_data[5, ]
```

* Looking at a set of rows and/or columns.

Here's all the entries in the 5th through 10th rows, 5th through 7th columns.
**NOTE: we use the `:` operator to look at a range of value.**

```{r}
trait_data[5:10, 5:7]
```

* For `data.frame`s, if you do not use row, column notation, you will get only the columns back.

```{r}
head(trait_data[5:7])
```

***

#### Challenge

What am I going to get if I execute the command below?

```{r, results='hide'}
head(trait_data[c("species","min_height")])
```

***

## Looking for missing elements

Many analyses in quantitative community ecology rely on multivariate statistical approaches, such as PCA, RDA, and NMDS. 
These analyses require complete data sets, and will not work if there are missing values. 
Below we go through an example of how to search for and replace a missing value. 

> **NOTE:** data imputation (i.e., replacing missing values with approximate values) is an area of study all by itself and we will not be covering it here. We encourage you to look in to this topic on your own, but here is a primer that might be useful to start with -- [https://libguides.princeton.edu/R-Missingdata](https://libguides.princeton.edu/R-Missingdata).


```{r missing-data}
## create matrix object
m <- matrix(c(1:8,NA), nrow=3, ncol=3) # a 3x3 matrix

## identify missing values
anyNA(m)
is.na(m[1,1])
is.na(m[2,2])
is.na(m[3,3])
is.na(m)

na_matrix <- is.na(df)
na_count <- sum(is.na(df))

## replace missing values with the average of all values
m[is.na(m)] <- mean(m, na.rm = TRUE)        # assign one value to any missing values

```


## Basic data visualization

There are many data visualization tools in R, and a very rich ecosystem of add-on [packages](https://cran.r-project.org/web/packages/index.html) that make it easy to create publication ready figures.
Here we will learn a few basic visualization tools.

### Visualization using the `ggplot2` package

We are going to introduce a data visualization package called **ggplot2**.
This package is great for producing publication quality graphics, but the syntax
used to create a plot is a little more involved than base R (i.e., the `graphics` package).

### Aside: Installing and loading packages

First, we need to install the `ggplot2` package.

**STOP** did you install the `tidyverse` package?

```{r installggplot2, eval=FALSE}
# Only need to do this once
install.packages("ggplot2")
```

Then load it:

```{r loadggplot2, message=FALSE}
library(ggplot2)
```

**NOTE: You will only have to install a new package once, but you will need to call the `library` function at the beginning of every new R session.**

### Histograms

Let's make a histogram of minimum heights in the trait data set.

```{r}
ggplot(data = trait_data, aes(x = min_height)) +
  geom_histogram()
```

Let's break down this call to introduce a few key things about ggplot

* ggplot: the initial canvas we're working on
* geom: geometric objects (i.e. the type of plot - histogram, points, line, etc) 
* aes: aesthetic mapping 

***

#### Challenge - use `?geom_histogram` to determine how to change the number of bins used

```{r, echo=FALSE, eval=FALSE}
ggplot(data = trait_data, aes(x = min_height)) +
  geom_histogram(bins = 15)
```

***

### Scatter plots

Next let's make a scatter plot, showing min_height versus max_height.

```{r}
ggplot(data = trait_data, aes(x = min_height, y = max_height)) +
  geom_point()
```

**THAT SEEMS SO COMPLICATED!**

It's true. The syntax for ggplot can seem pretty complicated. 
But the power of ggplot lies in the ability to lay several geometries (geoms) over each other.
Also, each geometry has a rich set of options. 
For example, let's add a trend line to the data we just plotted.

```{r}
ggplot(data = trait_data, aes(x = min_height, y = max_height)) +
  geom_point() +
  geom_smooth(method = "lm")
```

### Boxplots

Lastly, let's make a box plot. **NOTE:** we need to treat `perennial` as a factor, which here I'm doing 'on the fly'. 

```{r}
ggplot(data = trait_data, aes(x = as.factor(perennial), y = max_height)) +
  geom_boxplot()
```


***

#### Challenge

Using R, read in a data file from your own research, look at the summary statistics for a few of the columns, and then make a plot of your own choosing.

***

## Clean up your work space

It's good practice to keep your R work space (i.e., your **Environment**) tidy. 
At this point in the lesson we can go ahed and remove all of the variables we created.

```{r}
rm(list = ls())
```

