# Diversity Measures {#diversity}

***How diverse is this site? many sites? region?***

Let’s examine the "diversity" of our sites and study area. 
Diversity can encompass any sort of variation, e.g., taxonomic, trait, or phylogenetic. 
Here we assume that `spe` is a species abundance matrix where rows = sites, columns = taxa, and values must not be negative or missing.
*Recall that we did a validity test on the `spe` data table in Section \@ref(data-transform), ensuring there weren't any `NA` values. We could do a similar test looking for negative values.*

## Gamma (regional) diversity

```{r div-01, eval = TRUE, echo = TRUE}
gamma <- sum(colSums(spe) > 0)
gamma
```

## Alpha (per-site) diversity

```{r div-02, eval = TRUE, echo = TRUE}
### Alpha (per-site) diversity
alpha <- rowSums(spe > 0)  # within-site
avgalpha <- mean(alpha) # average within-site

avgalpha
```

## Beta (among-site) diversity: Whittaker’s

The calculation below is for a classic Whittaker's beta diversity measure, but we've subtracted a value of 1 from the ratio of gamma and avgalpha. 
This is because if we didn't subtract 1 and there was *no turnover*, then beta would equal 1, which can be confusing.

```{r div-03, eval = TRUE, echo = TRUE}
### Beta (among-site) diversity: Whittaker's
beta <- gamma / avgalpha - 1
beta
```

## Beta diversity: dust bunny indices (DBI)

McCune and Root (2015) proposed two model-free measures of how strongly multivariate species data exhibit a "dust bunny" distribution, that is, departure from multivariate normality because of many zero abundances due to few species occurring in any given sample unit. 
In other words, the DBI measures how much of the community matrix consists of "dust" - i.e., rare species, low abundances, and empty cells. 
High values indicate that most species occur in few sites and/or at low relative abundance. 
Low values indicate broadly distributed and abundant taxa.
This information is important because many multivariate methods become harder to interpret, and potentially invalid, when matrices have high dust-bunny values (many rare species and zeros), which is common in community ecology datasets.


**Calculate the proportion of zeros in the matrix, independent of abundance**

```{r div-04, eval = TRUE, echo = TRUE}
propzero <- sum(spe < .Machine$double.eps) / prod(dim(spe))
cat('Proportion of zeros in matrix:', propzero, '\n')
```

**Calculate the dust-bunny index (DBI) for the matrix, including abundances**

The code below includes many steps in one line. Here's what's going on:

- `vegan::decostand(spe, method='max')` = max-standardize each species column, i.e., divides each column by its maximum value, so each species abundance has a maximum value of 1
- `as.matrix(...)` = treats the output of the max-stadnardization as a `matrix`
- `mean(as.matrix(...))` = calculates the overall average abundance for all species at all sites
- `1 - mean(...)` = results in an index where a value of 0 indicates a dense matrix, with few zeros, where most species are common and a value of 1 indicates a sparse matrix, with many zeros, where rare species dominate

```{r div-05, eval = TRUE, echo = TRUE}
dbi <- 1 - mean(as.matrix(vegan::decostand(spe, method='max')))
cat('Dust bunny index:', dbi, '\n')
```


## Beta-diversity: no-share sites

High species turnover (beta-diversity) can cause many site-pairs to share no species in common. 
What proportion of all pairs of sites have totally different sets of species?

```{r}
### how many site-pairs share no species in common?
z <- vegan::no.shared(spe)
propnoshare <- sum(z) / length(z)
cat('Proportion of no-share sites:', propnoshare, '\n')
```

## Key references

McCune, B., and H.T. Root. 2015. Origin of the dust bunny distribution in ecological community data. Plant Ecology 216(5): 645-656.


