# Ordination (unconstrained)

***How best to order sites by dissimilarities of community compositions?***


> **TL;DR:** Ordination orders sites according to community compositions. 
Ordination distances indicate compositional dissimilarities. 
Nearby points are more compositionally similar than distant points.

Unconstrained ordination is a set of techniques that summarize patterns in a species abundance matrix without using any environmental variables or prior hypotheses about what drives those patterns. 
The ordination arranges sites (and sometimes species) in a low-dimensional space so that sites with similar community composition occur close together and dissimilar sites occur farther apart.
In community ecology, methods such as NMDS, PCoA, and PCA are commonly used as unconstrained ordinations to explore major gradients in species composition and identify natural patterns in the data before relating them to environmental factors. 

The three major unconstrained ordination methods are:

1. **Non-metric Multidimensional Scaling** - a rank-based method that focuses on preserving the order of dissimilarities rather than their exact values. NMDS is particularly useful for ecological data because it can handle non-linear relationships and is robust to outliers.
2. **Principal Coordinates Analysis (PCoA)** - a distance-based method that can use any dissimilarity measure (e.g., Bray-Curtis, Jaccard). PCoA is more flexible than PCA and can handle non-Euclidean distances, making it suitable for ecological data that often violate the assumptions of linearity.
3. **Principal Components Analysis (PCA)** - a linear method that reduces the dimensionality of the data by finding new axes (principal components) that maximize variance. PCA is best suited for continuous data and assumes linear relationships among variables.

Below we briefly show how to perform these three ordination methods in R using the `vegan` package for NMDS and PCoA, and the base R function `prcomp` for PCA.
Note that we are using the dissimilarity matrix `D` calculated in Section \@ref(dissimilarities) - **Stepacross adjustment** for NMDS and PCoA, while PCA is performed directly on the species abundance matrix `spe`.

```{r}
m1 <- metaMDS(D, k=2, maxit=250, try=100, trymax=101, trace=0) # NMDS
m2 <- cmdscale(D, k=2, add=T)                                  # PCoA
m3 <- prcomp(spe)                                              # PCA
```

We can visualize these results using the `ordiplot` function in the `vegan` package for NMDS and PCoA, and the base R plotting functions for PCA.

```{r}
### Visualizing ordinations
# color vector for plotting
u <- get_palette()
u <- u[1:nrow(spe)]
# compare three kinds of ordination
par(mfrow=c(1,3), bty='l', las=1)
plot(m1$points, type='n', xlab='NMDS1', ylab='NMDS2')
text(m1$points, rownames(m1$points),cex=.8, col=u)
plot(m2$points, type='n', xlab='PCoA1', ylab='PCoA2')
text(m2$points, rownames(m2$points),cex=.8, col=u)
plot(m3$x, type='n', xlab='PCA1', ylab='PCA2')
text(m3$x, rownames(m3$x),cex=.8, col=u)
par(mfrow=c(1,1))
```

***What do we see here?***

All three ordinations identify a similar primary gradient in community composition, separating sites on the left from those on the right. 
NMDS and PCoA produce very similar configurations because both are based on ecological dissimilarities among sites, whereas PCA uses Euclidean distances on the transformed species matrix and displays a pronounced horseshoe pattern. 
The similarity among methods increases confidence that the dominant pattern is real, while the differences highlight the assumptions made by each ordination technique.

***

#### Challenge

Re-rerun the PCA using a Hellinger transformation. What differences do you see in the ordination results?

```{r, eval=FALSE}
m3_hell <- prcomp(decostand(spe, "hellinger"))
```

<!---
Instructor note:

Hellinger transformation reduces the influence of double zeros and dominant species, making Euclidean distances more appropriate for species abundance data. As a result, PCA on Hellinger-transformed data often recovers community gradients that closely resemble those obtained from Bray-Curtis-based ordinations such as NMDS and PCoA, while retaining the advantages of PCA, including orthogonal axes and variance partitioning.

All three methods reveal essentially the same ecological pattern. The similarity among the ordinations suggests that the dominant gradient in species composition is strong and robust to the choice of ordination method. Hellinger transformation appears to have made PCA much more consistent with the Bray-Curtis-based NMDS and PCoA solutions.
--->

***

### Adding environmental information

We can add environmental information to the ordination plots to explore potential relationships between community composition and environmental gradients by coloring the points by environmental variable values.
Below we do this using `ggplot2` to visualize the NMDS results with points colored by three environmental variables: `k`, `k2o`, and `sand`.

```{r}
# To use ggplot, we need to first make a data.frame with the NMDS scores and the environmental variables
nmds_scores <- as.data.frame(vegan::scores(m1))
nmds_scores$k    <- env$k
nmds_scores$k2o  <- env$k2o
nmds_scores$sand <- env$sand

ggplot(nmds_scores,
             aes(NMDS1, NMDS2, color = k, label = rownames(nmds_scores))) +
  geom_text() +
  scale_color_viridis_c() +
  coord_equal() +
  theme_bw()

ggplot(nmds_scores,
             aes(NMDS1, NMDS2, color = k2o, label = rownames(nmds_scores))) +
  geom_text() +
  scale_color_viridis_c() +
  coord_equal() +
  theme_bw()

ggplot(nmds_scores,
             aes(NMDS1, NMDS2, color = sand, label = rownames(nmds_scores))) +
  geom_text() +
  scale_color_viridis_c() +
  coord_equal() +
  theme_bw()

```

## Ordination goodness-of-fit

Ordination methods have intrinsic measures of fit (e.g., stress for NMDS). However, for all methods, a "good" ordination should have a decent correlation between the Euclidean inter-point ordination distances versus the original species dissimilarities used to construct the ordination.

```{r}
cor(dist(m1$points), D, method='kendall') # NMS almost always best
cor(dist(m2$points), D, method='kendall')
cor(dist(m3$x),      D, method='kendall')
# cor(dist(m3_hell$x),      D, method='kendall')
```

> **NOTE:** In NMDS *stress* measures how well the NMDS plot represents the original dissimilarities among sites. Lower stress means the distances on the ordination plot are a good representation of the ecological distances in the original data; higher stress means the plot is a poorer representation.

## Dimensionality selection in NMDS

For NMDS, the user must select the number of dimensions a priori. 
Yet, what is the "best" number of dimensions? 
We can examine how stress changes as we increment the number of dimensions, and select the dimensionality that balances sufficiently "low" stress with the fewest dimensions.

```{r}
# define screeplot function, running NMDS for varying 'k' dimensions
scree_nms <- function(D, k=5, ...) {
  stress <- rep(NA, k)
  for (i in 1:k) {
    cat('calculating', i, 'of', k, 'dimensions...\n')
    stress[i] <- metaMDS(D, k=i, trace=0, ...)$stress
  }
  plot(1:k, stress, main='', xlab='Dimension', ylab='Stress',
       ylim=c(0, max(stress)*1.05), pch=16, las=1, bty='l')
  lines(1:k, stress)
  abline(0.20, 0, col='red', lty = 2)
  data.matrix(stress)
}

scree <- scree_nms(D, k=5, trymax=10)
scree
```

## Key references

Kruskal, J.B. 1964. Multidimensional scaling by optimizing goodness-of-fit to a nonmetric hypothesis. Psychometrika 29: 1–28.

Minchin, P.R. 1987. An evaluation of relative robustness of techniques for ecological ordinations. Vegetatio 69: 89–107.



