# Ordination (constrained)

**How best to order sites by dissimilarities of community compositions, in response to measured environmental variables?**

Constrained ordination addresses specific *a priori* hypotheses about variation in community compositions *attributable to measured environmental gradients*. 
It portrays only a fragment of all community variation, and is not used for exploratory analysis.

## Constrained ordination overview

Important and common constrained ordination tools include redundancy analysis (RDA; constrained PCA), canonical correspondence analysis (CCA), distance-based RDA (dbRDA, McArdle and Anderson 2001), and fuzzy set ordination (FSO, Roberts 2008). 
dbRDA and FSO both let the user select their own dissimilarity measure. 
Other methods (redundancy analysis or constrained correspondence analysis) force an implicit dissimilarity measure and make linearity/distributional assumptions.
The table below shows connections/analogs between unconstrained and constrained ordination methods.

| Method                | Unconstrained | Constrained |
| --------------------- | ------------- | ----------- |
| Linear / Euclidean    | PCA           | RDA         |
| Unimodal / Chi-square | CA            | CCA         |
| Distance-based        | PCoA          | dbRDA       |

## `dbRDA` -- the community analysis Swiss army knife

The dbRDA method, implemented in `vegan` with the `dbrda` function, is very versatile and can yield all of the ordination analyses we've been discussing based on how it is parameterized and the distance matrix used.

```{r}
## Define multiple different dissimilarity matrices
D_euc  <- vegdist(spe, 'euc')  # Euclidean
D_chi  <- vegdist(spe, 'chi')  # Chi-sq
D_bc   <- vegdist(spe, 'bray') # Bray-Curtis (Sorensen)

### PCA: PCO based on Euclidean distances (see `stats::prcomp`)
pca <- dbrda(D_euc ~ 1)

### CA/RA (correspondence analysis): PCO based on Chi-sq distances (see `vegan::cca`)
ca  <- dbrda(D_chi ~ 1)

### PCoA: generalizes to any dissimilarity (see `labdsv::pco`)
pco <- dbrda(D_bc  ~ 1)

### RDA: constrained form of PCA (see `vegan::rda`)
rda <- dbrda(D_euc ~ k2o + mg, data = env)

### CCA: constrained form of CA (see `vegan::cca`)
cca <- dbrda(D_chi ~ k2o + mg, data = env)

### dbRDA: constrained form of PCO; `lingoes` makes so can handle neg eigenvalues
dbr <- dbrda(D_bc  ~ k2o + mg, data = env, add = 'lingoes')
```

Visualize all of the methods together.

```{r}
library(patchwork)

u <- 1:NROW(spe)

plot_ord <- function(ord, xlab, ylab) {

  sc <- as.data.frame(vegan::scores(ord, display = "sites"))

  names(sc)[1:2] <- c("Axis1", "Axis2")
  sc$id <- seq_len(nrow(sc))

  ggplot(sc, aes(Axis1, Axis2)) +
    geom_text(aes(label = id, color = id)) +
    scale_color_viridis_c() +
    coord_equal() +
    labs(x = xlab, y = ylab) +
    theme_bw() +
    theme(
      legend.position = "none"
    )
}

p1 <- plot_ord(pca, "PCA1",   "PCA2")
p2 <- plot_ord(ca,  "CA1",    "CA2")
p3 <- plot_ord(pco, "PCoA1",  "PCoA2")
p4 <- plot_ord(rda, "RDA1",   "RDA2")
p5 <- plot_ord(cca, "CCA1",   "CCA2")
p6 <- plot_ord(dbr, "dbRDA1", "dbRDA2")

(p1 | p2 | p3) /
(p4 | p5 | p6)
```

**What do we see here?**

The dominant ecological gradient is sufficiently strong that it is recovered regardless of ordination method. 
Unconstrained methods (PCA, CA, PCoA) and constrained methods (RDA, CCA, dbRDA) all reveal a similar separation of sites, indicating that the major pattern in community composition is robust and is strongly associated with the environmental variables included in the constrained analyses.

## Visualizing site X environment relationships with RDA and dbRDA 



```{r}
rda_mod <- rda(spe ~ k2o + mg, data = env)

plot(rda_mod, type = "n", scaling = 2)
points(rda_mod, display = "sites", pch = 16, scaling = 2)
text(rda_mod, display = "species", col = "blue", scaling = 2)
text(rda_mod, display = "bp", col = "red", scaling = 2)

plot(rda_mod, type = "n", scaling = 1)
points(rda_mod, display = "sites", pch = 16, scaling = 1)
text(rda_mod, display = "species", col = "blue", scaling = 1)
text(rda_mod, display = "bp", col = "red", scaling = 1)
```


```{r}
# site_scores <- vegan::scores(rda, display = "sites", scaling = 2)
# env_scores <- vegan::scores(rda, display = "bp", scaling = 2)



plot(rda, type = "n", scaling = 2)
points(rda, display = "sites", pch = 16, scaling = 2)
text(rda, display = "bp", col = "red", scaling = 2)
```



```{r}
plot(dbr, type = "n", scaling = 2)
points(dbr, display = "sites", pch = 16, scaling = 2)
text(dbr, display = "bp", col = "red", scaling = 2)
```

