# Group indicator species

**Which species are indicative of pre-defined groups?**

The Indicator Value (*IndVal*) is the product of a species’ fidelity and specificity to a group (Dufrêne and Legendre 1997).

## Using the `indval` function

Here we use `labdsv::indval` to examine indicators of single groups, but also use R package `indicspecies` to test whether species might indicate multiple groups.

First, let’s use the same groups we used for PERMANOVA: which species are indicative of the four groups defined by soil potassium? 
We can compare these "real" indicator species against a random expectation.

```{r}
### real groups -- same as in the Group Differences section
grp <- cut(env$k, quantile(env$k, probs=seq(0,1,by=0.25)), include.lowest=T,
               labels=c('lo','med','hi','veryhi'))   # group memberships
iv  <- labdsv::indval(spe, grp) # indicator species analysis for *real* groups
summary(iv)                     # IndVal observed

### random groups
rnd <- sample(grp, length(grp), replace=T) # define random groups by bootstrapping
ivr <- labdsv::indval(spe, rnd) # indicator species analysis for *random* groups
summary(ivr)                    # IndVal expected at random
```

**How many indicator species would you expect at random?**

```{r}
### null expectation setting alpha = 5%
ceiling( ncol(spe) * 0.05 )
```

> **NOTE:** If you defined groups by a clustering method using species information, do not use *IndVal* to find indicator species of those groups (that would be circular logic).

## Key references

Dufrêne, M. and Legendre, P. 1997. Species assemblages and indicator species: the need for a flexible asymmetrical approach. Ecological Monographs 67(3): 345–366.

de Cáceres, M., P. Legendre, and M. Moretti. 2010. Improving indicator species analysis by combining groups of sites. Oikos 119:1674–1684.


